Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-13 paper-1/manuscript.tex
TODOs
Added sec Methods:Notation

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\!R^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}T^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\!D^{#2}_{#3}}

@@

@@

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\T{}{}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast in $\T{}{}$ at the start of target trials, i.e., $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$ corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{}$ tended to respond more strongly to target stimuli.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{}{}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{d} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{e} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression

@@

To sum up, we showed in Figure \ref{fig4} that greater $\T[msc]{A}{}$ in the portion of the network responding earliest to a stimulus plays an important role in reducing the overall response. Here, we showed that the neurons responding earliest to B exhibit higher $\T[dev]{B}{}$ as a consequence of an increased response, and that this increase in TA was directly responsible for reducing the response size and TA levels of the network as a whole.

@@

As noted previously, STD alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks evaluated in the full model with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the \std{} sequence, and least in the \dev{} sequence.

@@

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated the number of spikes in \dev{} target trials with $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.

@@

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. In order to respect this notion while at the same time minimizing the influence of activity differences between sequences, we used the average response $\R{S}{j} = \frac{1}{3} \sum_{seq \in (std, dev, msc)} \R[seq]{S}{j}$ of a neuron $j$ to a given stimulus $S$ as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $\R{S}{j} k x_j^S$, where we ignored added depression due to repeated spikes within a trial and instead defined $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    \D{S}{i} = \sum_{j \in pre} \R{S}{j} k (1-x_j^S)
\end{equation}
Notice that $\D{}{}$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $\T{}{}$, albeit more approximate.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target \dev{} trials, relative to \msc{}, whereas differences between target \dev{} and \msc{} trials are not a result of STD directly. To show this, we calculated $\D{}{}$ for A and B trials, and contrasted it across \dev{} and \msc{} sequences as with $\T{}{}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both sequences. The \msc{} sequence, however, clearly evoked greater depression, as borne out in the contrast $\Delta \D{A}{i}$, which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both sequences. In the \msc{} sequence, we can attribute the difference between A and B primarily to different response patterns: The context, and thus $x$, was approximately the same, but responses to B tended not to spread out as much in this network. However, in the \dev{} sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly spatially limited. As a result, the contrast $\Delta \D{B}{i}$ was slightly reduced in the periphery, but strikingly elevated near the stimulation site, indicating greatly increased synaptic depression in \dev{}.

@@

Finally, given the very clear contrast in $\D{A}{i}$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the \msc{} sequence, contrary to our stated hypothesis. If so, we should expect to see less $\D{A}{}$ in \dev{} than \msc{}. To confirm this, we analyzed $\Delta \D{A}{}$ across networks and stimuli, and found it to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120), as can be seen in the inset histogram in Figure \ref{fig8}c. Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in \msc{} than \dev{} in terms of $\D{}{}$, this difference appears to manifest only weakly in activity differences.

>>>>

% Common notation
\renewcommand{\R}[3][]{{}^{#1}_{}\!R^{#2}_{#3}}
\renewcommand{\T}[3][]{{}^{#1}_{}T^{#2}_{#3}}
\newcommand{\D}[3][]{{}^{#1}_{}\!D^{#2}_{#3}}
\newcommand{\mean}[1]{\overline{#1}}

@@

\subsection{Notation}\label{sec-notation}

Throughout this paper, we refer to three main quantities of interest: the stimulus response $\R{}{}$ defined below, the threshold adaptation value $\T{}{}$ defined in Equation \ref{eqn-TA}, and the short-term depression effect $\D{}{}$, which is introduced later in Equation \ref{eqn-D}. We employ sub- and superscripts to identify the precise quantity as follows, using $\R{}{}$ for illustration: $\R[seq]{X}{p}$ refers to the response of neuron(s) $p$, averaged across all trials of type $X$ in the sequence $seq$. Where $p$ does not identify individual neurons (e.g., $i$ or $j$), but rather a subpopulation (e.g., $exc$) or the whole network (no index), this refers to the median (of the individual quantity $\R[seq]{X}{i}$) across the indicated set of neurons. The trial type $X$ is either $A$ or $B$, referring to trials with stimulation at these sites, or $N$, indicating all \textbf{n}on-target trials B through E. Finally, we use the notation $\Delta \R{}{}$ to refer to the sequence contrast $\R[dev]{}{} - \R[msc]{}{}$, and $\mean{\R{}{p}}$ to indicate the mean, rather than median, of the individual quantities in a population $p$.

We define the response $\R{X}{i}$ as the mean spike count of neuron $i$ across time bins in trials of type $X$. Almost equivalently, we define $\T{X}{i}$ (and $\D{}{}$ analogously) as the mean across trials of type $X$ of $\T{}{i}(t=0)$, i.e., the average TA voltage at the beginning of trials, immediately before stimulation.

@@

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\T{}{}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast in $\T{}{}$ at the start of target trials, i.e., $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$ corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{}$ tended to respond more strongly to target stimuli.
% TODO: edit final sentence to end with ... tended to show a greater increase in their response to target stimuli.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{}{}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
% TODO: Drop "for this .. figures" since they're everywhere now.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{d} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{e} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression
% TODO: Amend to correctly identify early in panel b as the first 100 EXC. Consider making this consistent across sample and stats; currently, stats is indiscriminate.

@@

To sum up, we showed in Figure \ref{fig4} that greater $\T[msc]{A}{}$ in the portion of the network responding earliest to a stimulus plays an important role in reducing the overall response. Here, we showed that the neurons responding earliest to B exhibit higher $\T[dev]{B}{}$ as a consequence of an increased response, and that this increase in TA was directly responsible for reducing the response size and TA levels of the network as a whole.
% TODO replace with: To sum up, we showed in Figure \ref{fig4} that lower $\T[dev]{A}{}$ in the portion of the network responding earliest to A played an important role in increasing the overall response to A. Here, we showed that the neurons responding earliest to B exhibited higher $\T[dev]{B}{}$, which analogously reduced the response of the network as a whole to B. We also showed that this increase in $\T[dev]{B}{early}$ was a consequence of an increased response to non-target stimulation in the early portion, or in other words, of the more frequent presentation of B.

@@

As noted previously, STD alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks evaluated in the full model with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the \std{} sequence, and least in the \dev{} sequence.
% TODO: Show this, somewhere?

@@

To show in the sample network that the model does not maintain a history beyond the most recent trial, we correlated the number of spikes in \dev{} target trials with $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was strong and very clearly positive in the trial immediately following target (Pearson's $\rho$ = -0.984, p = 0; Wald test, n = 800), it was almost completely lost by the next trial ($\rho$ = -0.0525, p = 0.138) under the combined influence of decay and unrelated activity in the intervening trial.
% TODO: SFig for this.

@@

A measure of STD is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so it must be calculated based on observed activity. In order to respect this notion while at the same time minimizing the influence of activity differences between sequences, we used the average response $\R{S}{j} = \frac{1}{3} \sum_{seq \in (std, dev, msc)} \R[seq]{S}{j}$ of a neuron $j$ to a given stimulus $S$ as a measure of the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at $V_m = V_{rest}$ (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, we estimated the average peak EPSP evoked by a spike in presynaptic neuron $j$ as $\R{S}{j} k x_j^S$, where we ignored added depression due to repeated spikes within a trial and instead defined $x_j^S$ as the mean depression at the start of trials. Finally, summing over presynaptic neurons, we estimated the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    \D{S}{i} = \sum_{j \in pre} \R{S}{j} k (1-x_j^S) \label{eqn-D}
\end{equation}
Notice that $\D{}{}$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $\T{}{}$, albeit more approximate.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target \dev{} trials, relative to \msc{}, whereas differences between target \dev{} and \msc{} trials are not a result of STD directly. To show this, we calculated $\D{}{}$ for A and B trials, and contrasted it across \dev{} and \msc{} sequences as with $\T{}{}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both sequences. The \msc{} sequence, however, clearly evoked greater depression, as borne out in the contrast $\Delta \D{A}{i}$, which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both sequences. In the \msc{} sequence, we can attribute the difference between A and B primarily to different response patterns: The context, and thus $x$, was approximately the same, but responses to B tended not to spread out as much in this network. However, in the \dev{} sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly spatially limited. As a result, the contrast $\Delta \D{B}{i}$ was slightly reduced in the periphery, but strikingly elevated near the stimulation site, indicating greatly increased synaptic depression in \dev{}.
% TODO: After "different response patterns", clarify that these are caused by structural, not dynamic, differences

@@

Finally, given the very clear contrast in $\D{A}{i}$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the \msc{} sequence, contrary to our stated hypothesis. If so, we should expect to see less $\D{A}{}$ in \dev{} than \msc{}. To confirm this, we analyzed $\Delta \D{A}{}$ across networks and stimuli, and found it to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120), as can be seen in the inset histogram in Figure \ref{fig8}c. Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in \msc{} than \dev{} in terms of $\D{}{}$, this difference appears to manifest only weakly in activity differences.
% TODO: In hindsight, I'm not sure where this "weakly" claim comes from. Perhaps from Fig7, lower panels? If so, look at the scales in b and f and reconsider. Note also how the next paragraph seems to require a "not weak" conclusion here.
